---
title: "Assay Performance Report"
author:
  - name: Guthridge Informatics
    affiliation:
      - name: Oklahoma Medical Research Foundation
        department: Arthritis and Clinical Immunology
        city: Oklahoma City
        state: OK
        url: www.omrf.org
date: today
format:
  pdf:
    toc: true
editor:
  markdown:
    wrap: 72
params:
  project_name: "Example Serum"
  correction_procedure: "Plate Control Batch Correction"
  olink_assay_file: NULL
---

```{r}
#| label: load-files
#| include: false
#| echo: false
#| warning: false
#| message: false
#| results: hide
library(OlinkAssay)
oa <- readRDS(params$olink_assay_file)
# use later, when we finish that function. For now, use readRDS
# oa <- OlinkAssay::loadFromDisk()
software_version <- unique(S4Vectors::metadata(oa)[["SoftwareVersion"]])[[1]]
```

## Project Information

`r params$project_name` Olink Explore HT `r Sys.Date() `

`r params$correction_procedure` procedure used

**`r {stringr::str_glue("{R.version$version.string} used for QC and QC report")} `**

**`r {stringr::str_glue("Olink Explore HT software version {software_version} used for QC and QC report")} `**

## Manufacturer's Assay and Sample QC Information

The following graphs summarize the overall assay performance and per-assay
per-sample performance of the Olink Explore HT runs within the project. WARN and
FAIL protein assay are removed from the Level-2 QC process. Conservatively,
samples with any fail within any per-sample assay should be excluded from
subsequent analysis. These samples are kept throughout the Level-2 QC process,
however, they are labeled as "FAIL" in SampleQC column of the final Level-2
Parquet file. There are amplification, incubation, and extension controls that
should be removed prior to analysis as these are assay controls

{{< pagebreak >}}

```{r}
#| label: plot-qc-metrics
#| echo: false
#| warning: false
#| message: false
#| results: hide
#| fig-keep: all
# Basic piechart
QCPlot(oa)
```

```{r}
#| label: get-total-of-sample-per-each-assay
#| echo: false
#| warning: false
#| message: false

# bound_level1 <-
#   level_1_files |>
#   purrr::list_rbind()

sample_qc <- retrieveSampleQC(oa)
assay_qc <- retrieveAssayQC(oa)

n_assay <- dplyr::n_distinct(assay_qc[["OlinkID"]])
sample_tot <- dplyr::n_distinct(sample_qc[["SampleID"]])

sample_fails <-
  sample_qc |>
  dplyr::filter(SampleQC == "FAIL") |>
  dplyr::distinct(SampleID)

control_fails <-
  purrr::map(
    .x = c("negativeControls", "plateControls"),
    .f = \(x) {
      methods::slot(oa, x) |>
        tibble::as_tibble() |>
        dplyr::filter(SampleQC == "FAIL")
    }
  ) |>
  purrr::list_rbind() |>
  dplyr::distinct(SampleID, PlateID)

total_controls <-
  purrr::map(
    .x = c("negativeControls", "plateControls"),
    .f = \(x) {
      methods::slot(oa, x) |>
        tibble::as_tibble()
    }
  ) |>
  purrr::list_rbind() |>
  dplyr::distinct(SampleID, PlateID)

if (nrow(sample_fails > 0)) {
  message_sample <- stringr::str_glue(
    "Samples {glue::glue_collapse(sample_fails[['SampleID']], sep=', ')} were removed prior to level 2 normalization"
  )
} else {
  message_sample <- ""
}

if (nrow(control_fails > 0)) {
  message_control <- stringr::str_glue(
    "Controls {glue::glue_collapse(control_fails[['SampleID']], sep=', ')} were removed prior to level 2 normalization"
  )
} else {
  message_control <- ""
}
```

**`r {paste0(nrow(sample_fails), "/", nrow(sample_tot), " samples were excluded from determination of level 2 correction due to the failure of one or more analytes")}`**

`r {message_sample}`

**`r {paste0(nrow(control_fails), "/", nrow(total_controls), " controls were removed from level 2 determination of assay_level_qc and sample_level_qc due to the failure of one or more analytes")}`**

`r {message_control}`

{{< pagebreak >}}

**Samples with failed analytes and percentages**

```{r}
#| label: failed-analytes-and-percentages
#| echo: false
#| warning: false
#| message: false

table_failed <- sample_qc |>
  dplyr::group_by(PlateID, SampleID, SampleQC) |>
  dplyr::summarise(
    Analyte_Failure_Pct = round(100 * (dplyr::n()) / n_assay, digits = 2),
    .groups = 'drop'
  ) |>
  dplyr::filter(SampleQC == "FAIL") |>
  dplyr::select(-SampleQC)

flextable::set_flextable_defaults(fonts_ignore = TRUE)
table_failed |>
  dplyr::mutate(last_val = dplyr::row_number() == max(dplyr::row_number())) |>
  flextable::flextable(
    col_keys = c("PlateID", "SampleID", "Analyte_Failure_Pct")
  ) |>
  flextable::set_table_properties(align = "center") |>
  flextable::merge_v(j = ~PlateID) |>
  flextable::hline(i = ~ last_val == FALSE, border = officer::fp_border()) |>
  flextable::fix_border_issues() |>
  flextable::autofit()
```

{{< pagebreak >}}
## Level 2 QC Information

Tabulated QC performance metrics after Level-1 and Level-2 processing per each
source plate within the project. Remaining counts are quantifiable.\

```{r}
#| label: prep-plate_qc-for-use
#| echo: false

plate_qc <-
  assay(oa, "SampleAssayQC") |>
  tibble::as_tibble(rownames = "Assay") |>
  tidyr::pivot_longer(
    cols = -Assay,
    names_to = "SampleID",
    values_to = "sample_level_qc"
  ) |>
  dplyr::left_join(
    y = dplyr::select(
      tibble::as_tibble(colData(oa), rownames = "SampleID"),
      SampleID,
      PlateID
    ),
    by = dplyr::join_by("SampleID")
  )

plate_qc <-
  assay(oa, "ExtNPX") |>
  tibble::as_tibble(rownames = "Assay") |>
  tidyr::pivot_longer(
    cols = -Assay,
    names_to = "SampleID",
    values_to = "ExtNPX"
  ) |>
  dplyr::right_join(
    plate_qc,
    by = dplyr::join_by(Assay == Assay, SampleID == SampleID)
  ) |>
  dplyr::distinct()
```

```{r}
#| label: plate-level-qc
#| echo: false
#| warning: false
#| message: false
plate_qc |>
  dplyr::count(PlateID, sample_level_qc, name = "count") |>
  dplyr::group_by(PlateID) |>
  dplyr::mutate(percentage = round(100 * count / sum(count), digits = 2)) |>
  dplyr::filter(sample_level_qc != "Pass") |>
  dplyr::mutate(last_val = dplyr::row_number() == max(dplyr::row_number())) |>
  flextable::flextable(
    col_keys = c("PlateID", "sample_level_qc", "count", "percentage")
  ) |>
  flextable::set_table_properties(align = "center") |>
  flextable::merge_v(j = ~PlateID) |>
  flextable::hline(i = ~ last_val == TRUE, border = officer::fp_border()) |>
  flextable::fix_border_issues() |>
  flextable::autofit()
```

The following histograms show the distribution of the percentage of analytes
below the lower limits of detection per sample per 96-well source plate:\



```{r}
#| label: visualizing-the-per-plate-data
#| echo: false
#| warning: false
#| message: false
#| results: hide
#| fig-keep: all

upper_lim <- plate_qc |>
  dplyr::filter(sample_level_qc == "Below LLOD") |>
  dplyr::count(SampleID, sample_level_qc, name = "count") |>
  dplyr::mutate(lim_percentage = (100 * count / n_assay) + 10)

ylimit <- max(round(upper_lim[["lim_percentage"]], digits = 0))

p <- plate_qc |>
  dplyr::filter(sample_level_qc == "Below LLOD") |>
  dplyr::add_count(SampleID, sample_level_qc, name = "count") |>
  dplyr::mutate(percentage = (count / n_assay) * 100) |>
  dplyr::select(SampleID, percentage, PlateID) |>
  dplyr::distinct() |>
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = SampleID,
      y = percentage
    )
  ) +
  ggplot2::geom_col() +
  ggplot2::theme_classic() +
  ggplot2::ylab("% Total Assays Below LLOD") +
  ggplot2::theme(
    axis.text.x = ggplot2::element_blank(),
    axis.ticks.x = ggplot2::element_blank()
  ) +
  ggforce::facet_wrap_paginate(
    facets = ggplot2::vars(PlateID),
    ncol = 1,
    nrow = 4,
    page = 1
  ) +
  ggplot2::ylim(c(0, ylimit))

purrr::walk(
  .x = seq(ggforce::n_pages(p)),
  .f = \(plate_number) {
    print(
      p +
        ggforce::facet_wrap_paginate(
          facets = ggplot2::vars(PlateID),
          ncol = 1,
          nrow = 4,
          page = plate_number
        )
    )
  }
)
```

## Comparison of batch-corrected data to non-batch corrected

Batch-correction done using reference median calculated based on all source
plate Plate Controls.

### With High-Variance Analytes Retained

```{r}
#| label: analytes-with-failure-across-plate
#| echo: false
#| warning: false
#| message: false

missing_analytes <- plate_qc |>
  dplyr::filter(is.na(ExtNPX)) |>
  dplyr::count(Assay, name = "count") |>
  dplyr::filter(count > 80)

purrr::map2(
  .x = c("ExtNPX", "ExtNPX_Corrected", "LogProtExp_Raw"),
  .y = c("Raw ExtNPX", "Batch-Corrected ExtNPX", "Batch-Corrected LogProtExp"),
  .f = \(x, y) {
    batch_correction_umap(
      x = oa[which(!rownames(oa) %in% missing_analytes[["Assay"]]), ],
      assay_name = x,
      exclude_high_variance_assays = FALSE,
      title = y
    )
  }
)
```

### With High-Variance Analytes Removed

```{r}
#| label: remove-high-var-assays
#| echo: false
#| warning: false
#| message: false

purrr::map2(
  .x = c("ExtNPX", "ExtNPX_Corrected", "LogProtExp_Raw"),
  .y = c("Raw ExtNPX", "Batch-Corrected ExtNPX", "Batch-Corrected LogProtExp"),
  .f = \(x, y) {
    batch_correction_umap(
      x = oa,
      assay_name = x,
      exclude_high_variance_assays = TRUE,
      title = y
    )
  }
)
```

```{r}
#| label: get_high_var_assays
#| echo: false
#| warning: false
#| message: false

assay_exclude <-
  rowData(oa) |>
  tibble::as_tibble(rownames = "Assay") |>
  dplyr::mutate(
    high_var_assay = tidyr::replace_na(
      data = high_var_assay,
      replace = "Unknown"
    )
  ) |>
  dplyr::filter(high_var_assay == "High Variance")
```

**`r {nrow(assay_exclude)}`** analytes were removed from the UMAP due to C.V. >
20% for replicate Plate Controls. These technical high-variance analytes can be
identified by filtering high_var_assay for "High Variance" analytes.
