---
title: "Assay Performance Report"
author:
  - name: Guthridge Informatics
    affiliation:
      - name: Oklahoma Medical Research Foundation
        department: Arthritis and Clinical Immunology
        city: Oklahoma City
        state: OK
        url: www.omrf.org
date: today
format:
  pdf:
    toc: true
editor:
  markdown:
    wrap: 72
params:
  project_name: "Example Serum"
  correction_procedure: "Plate Control Batch Correction"
  level_1_file_path: NULL
  level_2_file_path: NULL
---

```{r}
#| label: load-files
#| include: false

level_1_files <- purrr::map(
  .x = list.files(
    path = params[["level_1_file_path"]],
    pattern = "parquet",
    full.names = TRUE
  ),
  .f = arrow::read_parquet,
  .progress = TRUE
)

level_2_files <-
  arrow::read_parquet(
    list.files(
      params[["level_2_file_path"]],
      pattern = "parquet",
      full.names = TRUE
    )
  )
```

## Project Information

`r params$project_name` Olink Explore HT `r Sys.Date() `

`r params$correction_procedure` procedure utilized

**`r {stringr::str_glue("{R.version$version.string} used for QC and QC report")} `**

**`r {stringr::str_glue("Olink Explore HT software version {unique(level_2_files$SoftwareVersion)} used for QC and QC report")} `**

## Manufacturer's Assay and Sample QC Information

The following graphs summarize the overall assay performance and per-assay
per-sample performance of the Olink Explore HT runs within the project. WARN and
FAIL protein assay are removed from the Level-2 QC process. Conservatively,
samples with any fail within any per-sample assay should be excluded from
subsequent analysis. These samples are kept throughout the Level-2 QC process,
however, they are labeled as "FAIL" in SampleQC column of the final Level-2
Parquet file. There are amplification, incubation, and extension controls that
should be removed prior to analysis as these are assay controls

{{< pagebreak >}}

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: hide
#| fig-keep: all

qc_check <- purrr::map(level_1_files, olinkqc::Olink_qc, .progress = TRUE)

# Basic piechart
purrr::map(qc_check, olinkqc::plot_qc)
```

```{r}
#| label: get-total-of-sample-per-each-assay
#| echo: false
#| warning: false
#| message: false

bound_level1 <-
  level_1_files |>
  purrr::list_rbind()

n_assay <-
  bound_level1 |>
  dplyr::count(OlinkID)

sample_fails <-
  bound_level1 |>
  dplyr::filter(
    SampleType == "SAMPLE",
    SampleQC == "FAIL"
  ) |>
  dplyr::distinct(SampleID)

sample_tot <-
  bound_level1 |>
  dplyr::filter(SampleType == "SAMPLE") |>
  dplyr::distinct(SampleID)

# control failure
control_fails <-
  bound_level1 |>
  dplyr::filter(
    SampleType != "SAMPLE",
    SampleQC == "FAIL"
  ) |>
  dplyr::distinct(PlateID, SampleID, WellID)

control_tot <-
  bound_level1 |>
  dplyr::filter(SampleType != "SAMPLE") |>
  dplyr::distinct(PlateID, SampleID, WellID)

if (nrow(sample_fails > 0)) {
  message_sample <- stringr::str_glue(
    "Samples {glue::glue_collapse(sample_fails[['SampleID']], sep=', ')} were removed prior to level 2 normalization"
  )
} else {
  message_sample <- ""
}

if (nrow(control_fails > 0)) {
  message_control <- stringr::str_glue(
    "Controls {glue::glue_collapse(control_fails[['SampleID']], sep=', ')} were removed prior to level 2 normalization"
  )
} else {
  message_control <- ""
}
```

**`r {paste0(nrow(sample_fails), "/", nrow(sample_tot), " samples were excluded from determination of level 2 correction due to the failure of one or more analytes")}`**

`r {message_sample}`

**`r {paste0(nrow(control_fails), "/", nrow(control_tot), " controls were removed from level 2 determination of assay_level_qc and sample_level_qc due to the failure of one or more analytes")}`**

`r {message_control}`

{{< pagebreak >}}

**Samples with failed analytes and percentages**

```{r}
#| label: failed-analytes-and-percentages
#| echo: false
#| warning: false
#| message: false

table_failed <- bound_level1 |>
  dplyr::group_by(PlateID, SampleID, SampleQC) |>
  dplyr::summarise(
    Analyte_Failure_Pct = round(100 * (dplyr::n()) / 5444, digits = 2),
    .groups = 'drop'
  ) |>
  dplyr::filter(SampleQC == "FAIL") |>
  dplyr::select(-SampleQC)

flextable::set_flextable_defaults(fonts_ignore = TRUE)
table_failed |>
  dplyr::mutate(last_val = dplyr::row_number() == max(dplyr::row_number())) |>
  flextable::flextable(
    col_keys = c("PlateID", "SampleID", "Analyte_Failure_Pct")
  ) |>
  flextable::set_table_properties(align = "center") |>
  flextable::merge_v(j = ~PlateID) |>
  flextable::hline(i = ~ last_val == FALSE, border = officer::fp_border()) |>
  flextable::fix_border_issues() |>
  flextable::autofit()
```

{{< pagebreak >}}
## Level 2 QC Information

Tabulated QC performance metrics after Level-1 and Level-2 processing per each
source plate within the project. Remaining counts are quantifiable.\

```{r}
#| label: plate-level-qc
#| echo: false
#| warning: false
#| message: false

plate_qc <- level_2_files |>
  dplyr::count(PlateID, sample_level_qc, name = "count") |>
  dplyr::group_by(PlateID) |>
  dplyr::mutate(percentage = round(100 * count / sum(count), digits = 2)) |>
  dplyr::filter(sample_level_qc != "Pass")

flextable::set_flextable_defaults(fonts_ignore = TRUE)

plate_qc |>
  dplyr::mutate(last_val = dplyr::row_number() == max(dplyr::row_number())) |>
  flextable::flextable(
    col_keys = c("PlateID", "sample_level_qc", "count", "percentage")
  ) |>
  flextable::set_table_properties(align = "center") |>
  flextable::merge_v(j = ~PlateID) |>
  flextable::hline(i = ~ last_val == TRUE, border = officer::fp_border()) |>
  flextable::fix_border_issues() |>
  flextable::autofit()
```

The following histograms show the distribution of the percentage of analytes
below the lower limits of detection per sample per 96-well source plate:\

```{r}
#| label: visualizing-the-per-plate-data
#| echo: false
#| warning: false
#| message: false
#| results: hide
#| fig-keep: all

plate_list <- split(level_2_files, f = level_2_files[["PlateID"]])

graph_bars <- function(.data) {
  .data |> 
    dplyr::filter(sample_level_qc == "Below LLOD") |> 
    dplyr::count(SampleID, sample_level_qc, name="count") |> 
    dplyr::mutate(percentage = 100 * count / 5440) |> 
    ggplot2::ggplot(
      mapping=ggplot2::aes(
        x = SampleID,
        y = percentage
        )
    ) +
      ggplot2::geom_col() +
      ggplot2::theme_classic() +
      ggplot2::ylab("% Total Assays Below LLOD") +
      ggplot2::theme(
        axis.text.x = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank()
      )
}

upper_lim <- level_2_files |> 
  dplyr::filter(sample_level_qc == "Below LLOD") |> 
  dplyr::count(SampleID, sample_level_qc, name="count") |> 
  dplyr::mutate(lim_percentage = (100 * count / 5444) + 10)

ylimit <- max(round(upper_lim[["lim_percentage"]], digits = 0))

purrr::imap(
  .x = plate_list,
  .f = \(plate, plate_name) {
    graph_bars(plate) + 
      ggplot2::ggtitle(plate_name) + 
      ggplot2::ylim(c(0, ylimit))
  }
)
```

## Comparison of batch-corrected data to non-batch corrected

Batch-correction done using reference median calculated based on all source
plate Plate Controls.

### With High-Variance Analytes Retained

```{r}
#| label: analytes-with-failure-across-plate
#| echo: false
#| warning: false
#| message: false

missing_analytes <- level_2_files |> 
  dplyr::filter(AssayType == "assay") |> 
  dplyr::filter(is.na(ExtNPX)) |> 
  dplyr::count(OlinkID, name="count") |> 
  dplyr::filter(count > 80)

purrr::map2(
  .x = c("ExtNPX", "ExtNPX_Corrected", "LogProtExp_Raw"),
  .y = c("Raw ExtNPX", "Batch-Corrected ExtNPX", "Batch-Corrected LogProtExp"),
  .f = \(x, y) olinkqc::batch_correction_umap_prep(
    .data = dplyr::filter(level_2_files, !OlinkID %in% missing_analytes[["OlinkID"]]),
    .col = x,
    exclude_high_variance_assays = FALSE,
    title = y
  )
)
```

### With High-Variance Analytes Removed

```{r}
#| echo: false
#| warning: false
#| message: false

assay_exclude <- level_2_files |> 
  dplyr::filter(high_var_assay == "High Variance") |> 
  dplyr::group_by(OlinkID, Assay) |> 
  dplyr::summarise()

purrr::map2(
  .x = c("ExtNPX", "ExtNPX_Corrected", "LogProtExp_Raw"),
  .y = c("Raw ExtNPX", "Batch-Corrected ExtNPX", "Batch-Corrected LogProtExp"),
  .f = \(x, y) olinkqc::batch_correction_umap_prep(
    .data = level_2_files,
    .col = x,
    exclude_high_variance_assays = TRUE,
    title = y
  )
)
```

**`r {nrow(assay_exclude)}`** analytes were removed from the UMAP due to C.V. >
20% for replicate Plate Controls. These technical high-variance analytes can be
identified by filtering high_var_assay for "High Variance" analytes.
