---
title: "Assay Performance Report"
author: "Guthridge Informatics"
date: "`r Sys.Date()`"
output: pdf_document

params:
  project_name: "Example Serum"
  correction_procedure: "Plate Control Batch Correction"
  level_1_file_path: NULL
  level_2_file_path: NULL
---


```{r include= FALSE}
library(tidyverse)
library(arrow)
library(flextable)
library(here)


source("\\\\data/ADI/Phenotyping_Core/01_Instrumentation/Olink_HT/02_Analysis_Projects/Olink_HT_QC/assoc_functions/Functions.R")
source("\\\\data/ADI/Phenotyping_Core/01_Instrumentation/Olink_HT/02_Analysis_Projects/Olink_HT_QC/assoc_functions/Graphing Functions.R")


# level_1_files <- lapply(list.files(here::here(
#   "SDP",
#   "Level_1"),
#                                    pattern = "parquet",
#                                    full.names = TRUE), arrow::read_parquet)

level_1_files <- lapply(list.files(params$level_1_file_path,
                                   pattern = "parquet",
                                   full.names = TRUE), arrow::read_parquet)

# level_1_files <- lapply(params$level_1_file_path, 
#                         arrow::read_parquet)

# level_2_files <- arrow::read_parquet(here::here("SDP", 
# "Level_2", 
# "Level_2_SDP.parquet"))

level_2_files <- arrow::read_parquet(list.files(params$level_2_file_path,
                                   pattern = "parquet",
                                   full.names = TRUE))

```
## Project Information  
`r params$project_name` Olink Explore HT `r Sys.Date() `

`r params$correction_procedure` procedure utilized

**`r {stringr::str_glue("{R.version$version.string} used for QC and QC report")} `**

**`r {stringr::str_glue("Olink Explore HT software version {unique(level_2_files$SoftwareVersion)} used for QC and QC report")} `**


## Manufacturer's Assay and Sample QC Information

The following graphs summarize the overall assay performance and per-assay per-sample performance of the Olink Explore HT runs within the project. WARN and FAIL protein assay are removed from the Level-2 QC process. Conservatively, samples with any fail within any per-sample assay should be excluded from subsequent analysis. These samples are kept throughout the Level-2 QC process, however, they are labeled as "FAIL" in SampleQC column of the final Level-2 Parquet file. There are amplification, incubation, and extension controls that should be removed prior to analysis as these are assay controls

```{r echo = FALSE, warning=FALSE, message=FALSE, results = 'hide', fig.keep = 'all'}
qc_check <- lapply(level_1_files, Olink_QC)
# Basic piechart

lapply(qc_check, Plot_QC)


```






```{r echo = FALSE, warning=FALSE, message=FALSE}
# get total of sample per each assay
bound_level1 <-
  level_1_files %>% 
  bind_rows()

n_assay <-
  bound_level1 %>%
  group_by(OlinkID) %>% 
  summarise(count = n())

sample_fails <-   
  bound_level1 %>%
  filter(SampleType == "SAMPLE") %>% 
  filter(SampleQC == "FAIL") %>% 
  group_by(SampleID) %>% 
  summarise(., .groups = 'drop') 

sample_tot <-  
  bound_level1 %>%
  filter(SampleType == "SAMPLE") %>% 
  group_by(SampleID) %>% 
  summarise(., .groups = 'drop')

  
# control failure
control_fails <- bound_level1 %>% 
  filter(SampleType != "SAMPLE") %>% 
  filter(SampleQC == "FAIL") %>% 
  group_by(PlateID, SampleID, WellID) %>% 
  summarise(., .groups = 'drop')

control_tot <- bound_level1 %>% 
  filter(SampleType != "SAMPLE") %>% 
  group_by(PlateID,SampleID, WellID) %>% 
  summarise(., .groups = 'drop')


if(nrow(sample_fails > 0)){
  message_sample <- stringr::str_glue("Samples ", paste0(pull(sample_fails, SampleID), collapse = ', '), "  were removed prior to level 2 normalization")
} else{ 
  message_sample <- paste0("")
}

if(nrow(control_fails > 0)){
  message_control <- stringr::str_glue("Controls ", paste0(pull(control_fails, SampleID), collapse = ', '), " were removed prior to level 2 normalization")
} else{
  message_control <- paste0("")
}

```

**`r {paste0(nrow(sample_fails), "/", nrow(sample_tot), " samples were excluded from determination of level 2 correction due to the failure of one or more analytes")}`**

`r {message_sample}`

**`r {paste0(nrow(control_fails), "/", nrow(control_tot), " controls were removed from level 2 determination of assay_level_qc and sample_level_qc due to the failure of one or more analytes")}`**

`r {message_control}`

**Samples with failed analytes and percentages**
```{r echo = FALSE, warning=FALSE, message=FALSE}
table_failed <- bound_level1 %>% 
  group_by(PlateID, SampleID, SampleQC) %>% 
  summarise(Analyte_Failure_Pct = round(100*(n())/5444, digits = 2), 
            .groups = 'drop') %>% 
  filter(SampleQC == "FAIL") %>% 
  select(-SampleQC)

flextable(table_failed) %>% 
  set_table_properties(., align = "center") %>% 
  merge_v(j = ~ PlateID) %>% 
  hline(i = rle(cumsum(.$body$spans$columns[,1] ))$values) %>% 
  fix_border_issues() %>% 
  flextable::autofit()
```

\pagebreak
## Level 2 QC Information  
Tabulated QC performance metrics after Level-1 and Level-2 processing per each source plate within the project. Remaining counts are quantifiable.   
```{r echo = FALSE, warning=FALSE, message=FALSE}
# Plate level qc

plate_qc <- level_2_files %>% 
  select(sample_level_qc, PlateID, OlinkID) %>% 
  group_by(PlateID, sample_level_qc) %>% 
  summarise(count = n(), .groups = 'drop') %>% 
  group_by(PlateID) %>% 
  mutate(percentage = round(100*count/sum(count), digits = 2)) %>% 
  filter(sample_level_qc != "Pass")

set_flextable_defaults(fonts_ignore = T)
flextable(plate_qc) %>% 
  set_table_properties(., align = "center") %>% 
  merge_v(j = ~ PlateID) %>% 
  hline(i = rle(cumsum(.$body$spans$columns[,1] ))$values) %>% 
  fix_border_issues() %>% 
  autofit()
  

```

The following histograms show the distribution of the percentage of analytes below the lower limits of detection per sample per 96-well source plate:  

```{r echo = FALSE, warning=FALSE, message=FALSE, results='hide', fig.keep = 'all'}
## visualizing the per-plate data
plate_list <- split(level_2_files, f = level_2_files$PlateID)

graph_bars <- function(data){

 data %>%
    filter(sample_level_qc == "Below LLOD") %>%
    group_by(SampleID, sample_level_qc) %>%
    summarise(count = n(), .groups = 'drop') %>%
    mutate(percentage = 100*count/5440) %>%  
    ggplot(aes(x = SampleID, y =  percentage))+
    geom_col()+
    theme_classic()+
    ylab("% Total Assays Below LLOD")+
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())

}


upper_lim <- level_2_files %>%
  filter(sample_level_qc == "Below LLOD") %>%
  group_by(SampleID, sample_level_qc) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(lim_percentage = (100*count/5444) + 10)

ylimit <- max(round(upper_lim$lim_percentage, digits = 0))

purrr::imap(
        .x = plate_list,
        .f = \(x, i) {
         graph_bars(x) + ggtitle(names(plate_list[i])) + ylim(c(0, ylimit))



        }
      )

```

## Comparison of batch-corrected data to non-batch corrected

Batch-correction done using reference median calculated based on all source plate Plate Controls.

### With High-Variance Analytes Retained
```{r echo = FALSE, warning=FALSE, message=FALSE}
# analytes with failure across plate
missing_analytes <- level_2_files |> 
  dplyr::filter(AssayType == "assay") |> 
  dplyr::filter(is.na(ExtNPX)) |> 
  dplyr::group_by(OlinkID) |> 
  dplyr::summarise(count = n()) |> 
  dplyr::filter(count > 80)

umap_prep <- function(column){
  
  
 pca_prep <- level_2_files |>
  dplyr::filter(
    SampleType == "SAMPLE", 
    Project != "Bridge"
  ) |>
  dplyr::filter(
    AssayType == "assay", 
    !OlinkID %in% missing_analytes$OlinkID
  ) |>
  dplyr::mutate(
    PlateID = stringr::str_split_i(
      PlateID,
      pattern = "_",
      i = 1
    )
  ) |>
  dplyr::mutate(ProteinID = paste0(Assay, "-", OlinkID)) |>
  dplyr::select(
    UniqueID,
    SampleID,
    PlateID,
    WellID,
    {{column}},
    ProteinID 
  ) |>
  # dplyr::mutate(LogProtExp_Raw = batch_corrected + log2(1e5)) |>
  tidyr::pivot_wider(
    id_cols = c(
      UniqueID,
      SampleID,
      PlateID,
      WellID
    ),
    names_from = ProteinID,
    values_from = {{column}}
  ) |>
  na.omit()

set.seed(123)

umap_res <- uwot::umap(
  pca_prep[-c(1:4)],
  scale = T,
  min_dist = 0.4
)


scores <- as.data.frame(umap_res)
colnames(scores) <- c("UMAP1", "UMAP2")

scores$PlateID <- pca_prep$PlateID
scores$SampleID <- pca_prep$SampleID

scores |>
  ggplot(aes(UMAP1, UMAP2, color = PlateID)) +
  geom_point(size = 3) +
  theme_minimal()

}


umap_prep(ExtNPX)+
  ggtitle("Raw ExtNPX")+
  theme(legend.text = element_text(face = "bold", 
                                   size = 12), 
        legend.title = element_blank(), 
        axis.title = element_text(size = 12))

umap_prep(ExtNPX_Corrected)+
  ggtitle("Batch-Corrected ExtNPX")+
  theme(legend.text = element_text(face = "bold", 
                                   size = 12), 
        legend.title = element_blank(), 
        axis.title = element_text(size = 12))

umap_prep(LogProtExp_Raw)+
  ggtitle("Batch-Corrected LogProtExp")+
  theme(legend.text = element_text(face = "bold", 
                                   size = 12), 
        legend.title = element_blank(), 
        axis.title = element_text(size = 12))

```


### With High-Variance Analytes Removed  

```{r echo = FALSE, warning = FALSE, message = FALSE}

assay_exclude <- level_2_files %>% 
  filter(high_var_assay == "High Variance") %>% 
  group_by(OlinkID, Assay) %>% 
  summarise()

umap_prep <- function(column){
  
  
 pca_prep <- level_2_files |>
  dplyr::filter(
    SampleType == "SAMPLE", 
    Project != "Bridge"
  ) |>
  dplyr::filter(
    AssayType == "assay", 
    !OlinkID %in% missing_analytes$OlinkID, 
    !OlinkID %in% assay_exclude$OlinkID
  ) |>
  dplyr::mutate(
    PlateID = stringr::str_split_i(
      PlateID,
      pattern = "_",
      i = 1
    )
  ) |>
  dplyr::mutate(ProteinID = paste0(Assay, "-", OlinkID)) |>
  dplyr::select(
    UniqueID,
    SampleID,
    PlateID,
    WellID,
    {{column}},
    ProteinID 
  ) |>
  # dplyr::mutate(LogProtExp_Raw = batch_corrected + log2(1e5)) |>
  tidyr::pivot_wider(
    id_cols = c(
      UniqueID,
      SampleID,
      PlateID,
      WellID
    ),
    names_from = ProteinID,
    values_from = {{column}}
  ) |>
  na.omit()

set.seed(123)

umap_res <- uwot::umap(
  pca_prep[-c(1:4)],
  scale = T,
  min_dist = 0.4
)


scores <- as.data.frame(umap_res)
colnames(scores) <- c("UMAP1", "UMAP2")

scores$PlateID <- pca_prep$PlateID
scores$SampleID <- pca_prep$SampleID

scores |>
  ggplot(aes(UMAP1, UMAP2, color = PlateID)) +
  geom_point(size = 3) +
  theme_minimal()

}


umap_prep(ExtNPX)+
  ggtitle("Raw ExtNPX")+
  theme(legend.text = element_text(face = "bold", 
                                   size = 12), 
        legend.title = element_blank(), 
        axis.title = element_text(size = 12))

umap_prep(ExtNPX_Corrected)+
  ggtitle("Batch-Corrected ExtNPX")+
  theme(legend.text = element_text(face = "bold", 
                                   size = 12), 
        legend.title = element_blank(), 
        axis.title = element_text(size = 12))

umap_prep(LogProtExp_Raw)+
  ggtitle("Batch-Corrected LogProtExp")+
  theme(legend.text = element_text(face = "bold", 
                                   size = 12), 
        legend.title = element_blank(), 
        axis.title = element_text(size = 12))


```

**`r {nrow(assay_exclude)}`** analytes were removed from the UMAP due to C.V. > 20% for replicate Plate Controls. These technical high-variance analytes can be identified by filtering high_var_assay for "High Variance" analytes.
